Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2024-10-17 manuscript_latex/manuscript.tex
Further manuscript preparations.
Manuscript main body
It compiles!

To do:
- add appendix
- replace citations

@@

%% Abstract
\begin{abstract}
%% Text of abstract
Abstract text.
\end{abstract}

@@

%%Research highlights
\begin{highlights}
\item Research highlight 1
\item Research highlight 2
\end{highlights}

%% Keywords
\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

@@

%% main text
%%

%% Use \section commands to start a section
\section{Example Section}
\label{sec1}
%% Labels are used to cross-reference an item using \ref command.

Section text. See Subsection \ref{subsec1}.

%% Use \subsection commands to start a subsection.
\subsection{Example Subsection}
\label{subsec1}

Subsection text.

%% Use \subsubsection, \paragraph, \subparagraph commands to
%% start 3rd, 4th and 5th level sections.
%% Refer following link for more details.
%% https://en.wikibooks.org/wiki/LaTeX/Document_Structure#Sectioning_commands

\subsubsection{Mathematics}
%% Inline mathematics is tagged between $ symbols.
This is an example for the symbol $\alpha$ tagged as inline mathematics.

%% Displayed equations can be tagged using various environments.
%% Single line equations can be tagged using the equation environment.
\begin{equation}
f(x) = (x+a)(x+b)
\end{equation}

%% Unnumbered equations are tagged using starred versions of the environment.
%% amsmath package needs to be loaded for the starred version of equation environment.
\begin{equation*}
f(x) = (x+a)(x+b)
\end{equation*}

%% align or eqnarray environments can be used for multi line equations.
%% & is used to mark alignment points in equations.
%% \\ is used to end a row in a multiline equation.
\begin{align}
 f(x) &= (x+a)(x+b) \\
      &= x^2 + (a+b)x + ab
\end{align}

\begin{eqnarray}
 f(x) &=& (x+a)(x+b) \nonumber\\ %% If equation numbering is not needed for a row use \nonumber.
      &=& x^2 + (a+b)x + ab
\end{eqnarray}

%% Unnumbered versions of align and eqnarray
\begin{align*}
 f(x) &= (x+a)(x+b) \\
      &= x^2 + (a+b)x + ab
\end{align*}

\begin{eqnarray*}
 f(x)&=& (x+a)(x+b) \\
     &=& x^2 + (a+b)x + ab
\end{eqnarray*}

%% Refer following link for more details.
%% https://en.wikibooks.org/wiki/LaTeX/Mathematics
%% https://en.wikibooks.org/wiki/LaTeX/Advanced_Mathematics

%% Use a table environment to create tables.
%% Refer following link for more details.
%% https://en.wikibooks.org/wiki/LaTeX/Tables
\begin{table}[t]%% placement specifier
%% Use tabular environment to tag the tabular data.
%% https://en.wikibooks.org/wiki/LaTeX/Tables#The_tabular_environment
\centering%% For centre alignment of tabular.
\begin{tabular}{l c r}%% Table column specifiers
%% Tabular cells are separated by &
  1 & 2 & 3 \\ %% A tabular row ends with \\
  4 & 5 & 6 \\
  7 & 8 & 9 \\
\end{tabular}
%% Use \caption command for table caption and label.
\caption{Table Caption}\label{fig1}
\end{table}

%% Use figure environment to create figures
%% Refer following link for more details.
%% https://en.wikibooks.org/wiki/LaTeX/Floats,_Figures_and_Captions
\begin{figure}[t]%% placement specifier
%% Use \includegraphics command to insert graphic files. Place graphics files in
%% working directory.
\centering%% For centre alignment of image.
\includegraphics{example-image-a}
%% Use \caption command for figure caption and label.
\caption{Figure Caption}\label{fig1}
%% https://en.wikibooks.org/wiki/LaTeX/Importing_Graphics#Importing_external_graphics
\end{figure}

>>>>

%% The booktabs package provides table makeup macros.
\usepackage{booktabs}

%% The hyperref package allows for typesetting hyperlinks.
%% \usepackage{hyperref}

%% Custom math operators
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}

@@

%% Abstract
% You are required to provide a concise and factual abstract which does not exceed
% 250 words. The abstract should briefly state the purpose of your research,
% principal results and major conclusions. Some guidelines:
%
% Abstracts must be able to stand alone as abstracts are often presented separately from the article.
%
% Avoid references. If any are essential to include, ensure that you cite the author(s) and year(s).
%
% Avoid non-standard or uncommon abbreviations. If any are essential to include,
% ensure they are defined within your abstract at first mention.

\begin{abstract}

\end{abstract}

@@

%%Research highlights
% Highlights are a short collection of bullet points that should capture the novel
% results of your research as well as any new methods used during your study.
% Highlights will help increase the discoverability of your article via search engines.
% Some guidelines:
%
% Submit highlights as a separate editable file in the online submission system
% with the word "highlights" included in the file name.
%
% Highlights should consist of 3 to 5 bullet points, each a maximum of 85 characters, including spaces.

\begin{highlights}
\item Research highlight 1
\item Research highlight 2
\end{highlights}

%% Keywords
% You are required to provide 1 to 7 keywords for indexing purposes. Keywords should be written in English.
% Please try to avoid keywords consisting of multiple words (using "and" or "of").
% We recommend that you only use abbreviations in keywords if they are firmly established in the field.
\begin{keyword}

%% keywords here, in the form: keyword \sep keyword

@@

%% main text
%%
\section{Introduction}\label{introduction}

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
(Bartlett 1947; John W. Tukey 1957). The two most commonly used
transformations are that of Box and Cox (1964) and Yeo and Johnson
(2000). The Box-Cox transformation of a feature value \(x_i\) of feature
\(\mathbf{X}=\left\{x_1, x_2, \ldots, x_n \right\}\) under the
transformation parameter \(\lambda\) is defined as:

\begin{equation}
\label{eqn:box-cox-original}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}

One limitation of the Box-Cox transformation is that it is only defined
for \(x_i > 0\). In contrast, the Yeo-Johnson transformation under the
transformation parameter \(\lambda\) is defined for any
\(x_i \in \mathbb{R}\):

\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}

The \(\lambda\)-parameter is typically optimised using maximum
likelihood estimation under the assumption that the transformed feature
is normally distributed. As noted by Raymaekers and Rousseeuw, this
approach is sensitive to outliers, and robust versions of Box-Cox and
Yeo-Johnson transformations were devised (Raymaekers and Rousseeuw
2024).

Applying a power transformation does not guarantee that transformed
features are normally distributed. Depending on location and scale of a
feature and the presence of outliers, power transformations may decrease
normality, as shown in Figure \ref{fig:decreased-normality}. If
normality of the transformed feature is not checked, e.g.~in automated
power transformation in machine learning workflows, several issues may
arise. For example, statistical tests such as ANOVA may produce
incorrect results due to violation of the normality assumption.
Likewise, machine learning methods that assume normality of input
features may suffer a decrease in performance. Moreover, large negative
or positive \(\lambda\)-parameters may lead to numeric issues in any
subsequent computations.

Statistical tests for normality, such as the Shapiro-Wilk test (Shapiro
and Wilk 1965), could be automatically applied to transformed features.
However, given sufficiently large sample sizes such tests can detect
trivial deviations from normality, and may lead to rejection of
sufficiently good power transformations.

\begin{figure}

{\centering \includegraphics{figure_1}

}

\caption{Effect of location, scale and outliers on estimation of the Box-Cox and Yeo-Johnson transformation parameter $\lambda$. $10000$ samples were drawn from a normal distribution: $\mathcal{N}(\mu, 1)$ for the shift dataset, $\mathcal{N}(10, \sigma)$ for the scale dataset and $\mathcal{N}(0, 1)$ for the outlier dataset. Additionally, an outlier with value $d$ was added to the outlier dataset. Since samples are drawn from a normal distribution, a transformation parameter of $\lambda = 1$ is expected. However, a large shift in location, a scale that is small compared to the location, or presence of large outliers lead to incorrectly estimated transformation parameter values.}\label{fig:decreased-normality}
\end{figure}

To address these issues, we make the following contributions:

\begin{itemize}
\item
  We devise location- and scale-invariant versions of the Box-Cox and
  Yeo-Johnson transformation, including versions robust to outliers.
\item
  We derive the maximum likelihood criterion for location- and
  scale-invariant Box-Cox and Yeo-Johnson transformations to allow for
  optimising transformation parameters.
\item
  We define an empirical central normality test for detecting cases
  where power transformations fail to yield an approximately normally
  distributed transformed feature.
\item
  We assess the effect of power transformations on the performance of
  machine learning models.
\end{itemize}

\section{Theory}\label{theory}

In this section, we will first introduce location- and scale-invariant
versions of the Box-Cox and Yeo-Johnson transformations. Subsequently,
we define weighted location- and scale-invariant transformations and
weighting methods for robust transformations. We then define the
quantile function for asymmetric generalised normal distributions to
enable random sampling. Finally, we define the overall framework for the
empirical central normality test.

\subsection{Location- and scale-invariant power
transformation}\label{location--and-scale-invariant-power-transformation}

Box-Cox and Yeo-Johnson transformations are modified by introducing
shift parameter \(x_0\) and scale parameter \(s\) into equations
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The
location- and scale-invariant Box-Cox transformation of a feature value
\(x_i\) of feature \(\mathbf{X}\) under transformation parameter
\(\lambda\), shift parameter \(x_0\) and scale parameter \(s\) is then:

\begin{equation}
\label{eqn:box-cox-invariant}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}

where \(x_i - x_0 > 0\). Likewise, the location- and scale-invariant
Yeo-Johnson transformation of a feature value \(x_i\) under
transformation parameter \(\lambda\), shift parameter \(x_0\) and scale
parameter \(s\) is:

\begin{equation}
\label{eqn:yeo-johnson-invariant}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in Appendix A.
The location- and scale-invariant Box-Cox log-likelihood function is:

\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

subject to \(x_i - x_0 > 0\). \(\mu\) and \(\sigma^2\) are the mean and
variance of the Box-Cox transformed feature
\(\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})\), respectively.
Similarly, the location- and scale-invariant Yeo-Johnson log-likelihood
function is:

\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}

where \(\mu\) and \(\sigma^2\) are the mean and variance of the
Yeo-Johnson transformed feature
\(\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})\), respectively.

\subsection{Robust location- and scale-invariant power
transformations}\label{robust-location--and-scale-invariant-power-transformations}

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by Raymaekers and Rousseeuw
(2024), the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. Raymaekers and Rousseeuw
(2024) achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to Raymaekers
and Rousseeuw (2024), these log-likelihood functions includes additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

\begin{equation}
\label{eqn:box-cox-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

where \(\mu_w\) and \(\sigma^2_w\) are the weighted mean and weighted
variance of the Box-Cox transformed feature
\(\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})\):

\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{BC}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}

Analogously, the weighted location- and scale-invariant Yeo-Johnson
log-likelihood function is:

\begin{equation}
\label{eqn:yeo-johnson-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}

where \(\mu_w\) and \(\sigma^2_w\) are the weighted mean and weighted
variance of the Yeo-Johnson transformed feature
\(\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})\):

\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{YJ}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}

The weights \(w_i\) in equations
\ref{eqn:box-cox-weighted-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-weighted-invariant-log-likelihood} can be set using
several weighting functions. Using \(\dot{x}_i\) as an argument that
will be defined later, we investigate three weighting functions:

\begin{itemize}
\item
  A step function, with \(\delta_1 \geq 0\) as threshold parameter:
\end{itemize}

\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| \leq \delta_1\\
0 & \text{if } \left| \dot{x}_i \right| > \delta_1
\end{cases}
\end{equation}

\begin{itemize}
\item
  A triangle function (or generalised Huber weight), with
  \(\delta_1 \geq 0\) and \(\delta_2 \geq \delta_1\) as threshold
  parameters:
\end{itemize}

\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
1 - \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
0 & \text{if } \left| \dot{x}_i \right| > \delta_2
\end{cases}
\end{equation}

\begin{itemize}
\item
  A tapered cosine function (J. W. Tukey 1967), with \(\delta_1 \geq 0\)
  and \(\delta_2 \geq \delta_1\) as threshold parameters:
\end{itemize}

\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
0.5 + 0.5 \cos\left(\pi \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} \right) & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
0 & \text{if } \left| \dot{x}_i \right| > \delta_2
\end{cases}
\end{equation}

All weighting functions share the characteristic that for
\(\left| \dot{x}_i \right|< \delta_1\), instances are fully weighted,
i.e.~when \(\delta_1 > 0\) the weighting functions are symmetric window
functions with a flat top. The triangle and tapered cosine functions
then gradually down-weight instances with
\(\delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2\), and assign no
weight to instances \(\left| x_i \right| > \delta_2\). Examples of these
weighting function are shown in Figure \ref{fig:weighting-functions}.

\begin{figure}

{\centering \includegraphics{figure_2}

}

\caption{Weighting functions investigated in this study to make power transformations more robust against outliers. In this example, the step function was parameterised with $\delta_1 = 0.60$. The triangle and tapered cosine functions were both parameterised with $\delta_1 = 0.30$ and $\delta_2 = 0.90$.}\label{fig:weighting-functions}
\end{figure}

Each weighting function has an argument \(\dot{x}\) that is related to
the (transformed) feature in one of several ways:

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After Raymaekers and
  Rousseeuw (2024),
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) (Huber 1981). Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

\subsection{Asymmetric generalised normal
distributions}\label{asymmetric-generalised-normal-distributions}

Modifications intended to make power transformations invariant to
location and scale of a feature and methods to improve their robustness
against outliers need to be assessed using data drawn from a range of
different distributions. Since the power transformations are intended
for use with unimodal distributions, the generalised normal distribution
(Subbotin 1923; Nadarajah 2005) is a suitable option for simulating
realistic feature distributions. This distribution has the following
probability density function \(f_{\beta}\) for a value
\(x \in \mathbb{R}\):

\begin{equation}
f_{\beta}(x) = \frac{\beta}{2\Gamma\left(1 / \beta \right)} e^{-\left| x \right|^\beta}
\end{equation}

Here, \(\Gamma\) is the gamma function, and \(\beta\) is a strictly
positive shape parameter. For \(\beta = 1\), the probability density
function describes a Laplace distribution. A normal distribution is
found for \(\beta=2\), and for large \(\beta\), the distribution
approaches a uniform distribution. We will refrain from introducing
scale and location parameters here directly.

Realistic feature distributions may be skewed. Gijbels et al.~describe a
recipe for introducing skewness into the otherwise symmetric generalised
normal distribution (Gijbels, Karim, and Verhasselt 2019), leading to
the following probability density function:

\begin{equation}
f_{\alpha}(x; \mu, \sigma, \beta) = \frac{2 \alpha \left(1 - \alpha\right)}{\sigma}
\begin{cases}
f_{\beta}\left( \left(1 - \alpha \right) \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x \leq \mu \\
f_{\beta}\left( \alpha \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x > \mu
\end{cases}
\end{equation}

Here, \(\alpha \in (0,1)\) is a skewness parameter. \(\alpha > 0.5\)
creates a distribution with a negative skew, i.e.~a left-skewed
distribution. A right-skewed distribution is created for
\(\alpha < 0.5\). \(\mu \in \mathcal{R}\) and \(\sigma \in (0, \infty)\)
are location and scale parameters, respectively. \(f_{\alpha}\) thus
describes the probability density function of an asymmetric generalised
normal distribution, which we will refer to here and parametrise as
\(\mathcal{AGN}\left(\mu, \sigma, \alpha, \beta \right)\).

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the following
quantile function \(F_{\alpha}^{-1}(p)\):

\begin{equation}
F_{\alpha}^{-1}(p; \mu, \sigma, \beta) =
\begin{cases}
\mu + \frac{\sigma}{1 - \alpha} F_{\beta}^{-1} \left( \frac{p}{2 \alpha}\right) & \text{, } p \leq \alpha \\
\mu + \frac{\sigma}{\alpha} F_{\beta}^{-1} \left( \frac{1 + p - 2 \alpha}{2 \left(1 - \alpha \right)} \right) & \text{, } p > \alpha
\end{cases}
\end{equation}

The quantile function for the asymmetric generalised normal distribution
\(F_{\alpha}^{-1}\) thus incorporates the quantile function
\(F_{\beta}^{-1}\) of the symmetric generalised normal distribution.
\(F_{\beta}^{-1}\) was derived by Griffin to be (Griffin 2018):

\begin{equation}
F_{\beta}^{-1}(p) = \sgn\left(p - 0.5 \right) F_{\Gamma}^{-1}\left(2 \left|p - 0.5 \right|; 1 / \beta \right)
\end{equation}

Here, \(F_{\Gamma}^{-1}\) is the quantile function of the gamma
distribution with shape \(1 / \beta\), which can be numerically
approximated.

\subsection{Empirical central normality
test}\label{empirical-central-normality-test}

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test (Shapiro and Wilk 1965). In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality. The null
hypothesis \(H_0\) is that the distribution is centrally normal. The
alternative hypothesis \(H_1\) is that the distribution is not centrally
normal.

Let central normality be defined as the normality of central portion
\(\kappa\) of the data,
i.e.~\(\mathbf{X}_{\text{central}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}\),
with \(p_i\) probabilities of the empirical distribution, as previously.
We then compute the residual errors between the z-scores of the
transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
expected z-scores from a standard normal distribution:
\(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
with \(\mu_M\) and \(\sigma_M\) robust Huber M-estimates of location and
scale of the transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\)
(Huber 1981). The set of residual errors for the central portion of the
data is then
\(\mathbf{R}_{\text{central}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}\).

The test statistic \(\tau_{\text{ecn}}\) is then defined as:

\begin{equation}
\tau_{\text{ecn}} = \frac{\sum_{i=1}^{N} r_i \left[r_i \in \mathbf{R}_{\text{central}}\right]}{\sum_{i=1}^N \left[r_i \in \mathbf{R}_{\text{central}}\right]} 
\end{equation}

Here \([\quad]\) denotes an Iverson bracket. The test statistic is equal
to the mean of the residual errors of the central portion of the data.

To use the test statistic, the central portion of the data needs to be
defined and the Type 1 error rates determined. We will do so in
section \ref{simulation}.

\section{Simulation}\label{simulation}

We used simulated data to assess invariance to location and scale of the
proposed power transformations, weighting for robust transformations,
and to develop the empirical central normality test. The \(\lambda\)
parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation (Powell
2009) through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
(Zwanenburg and Löck 2024b) (version 1.0.0). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

\subsection{Invariance to location and
scale}\label{invariance-to-location-and-scale}

To assess whether the proposed power transformations lead to values of
\(\lambda\) that are invariant to location and scale of the
distribution, we simulated three different distributions. We first
randomly drew \(10000\) values from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed generalised normal
distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed generalised normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).
We then computed transformation parameter \(\lambda\) using the original
definitions (Eqn. \ref{eqn:box-cox-original} and
\ref{eqn:yeo-johnson-original}) and the location- and scale-invariant
definitions (Eqn. \ref{eqn:box-cox-invariant} and
\ref{eqn:yeo-johnson-invariant}) for each distribution. To assess
location invariance, a positive value \(d_{\text{shift}}\) was added to
each distribution with \(d_{\text{shift}} \in [1, 10^6]\). Similarly, to
assess scale invariance, each distribution was multiplied by a positive
value \(d_{\text{scale}}\), where \(d_{\text{scale}} \in [1, 10^6]\).

\begin{figure}

{\centering \includegraphics{figure_3}

}

\caption{Invariant power transformation produces transformation parameters that are invariant to location and scale. Samples were drawn from normal, right-skewed and left-skewed distributions, respectively, which then underwent a shift $d_{\text{shift}}$ or multiplication by $d_{\text{scale}}$. Estimates of the transformation parameter $\lambda$ for the conventional power transformations show strong dependency on the overall location and scale of the distribution, whereas estimates obtained for the location- and scale-invariant power transformations are constant.}\label{fig:shifted-distributions}
\end{figure}

The result is shown in Figure \ref{fig:shifted-distributions}. For each
distribution, transformation parameter \(\lambda\) varied with
\(d_{\text{shift}}\) and \(d_{\text{scale}}\) when estimated for
conventional transformations. In contrast, estimation of \(\lambda\) for
invariant power transformations was invariant to both
\(d_{\text{shift}}\) and \(d_{\text{scale}}\).

\subsection{Robust transformations}\label{robust-transformations}

Outliers may be present in data and affect estimation of transformation
parameters. The log-likelihood function can be weighted to assign less
weight to outlier instances, see equations
\ref{eqn:box-cox-weighted-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-weighted-invariant-log-likelihood}. We proposed
three weighting function: step, triangle and tapered cosine, that have
one, two and two parameters, respectively. Each weighting function then
takes one of three values as input: probabilities of the empirical
distribution of the original feature, the z-score of the transformed
feature values, or the residual error between the z-score of the
transformed feature values and their expected z-score based on the
normal distribution.

To determine the weighting function parameters for each of the nine
combinations, \(m_d=100\) asymmetric generalised normal distributions
were drawn. Each distribution was parametrised with a randomly chosen
skewness parameter \(\alpha \sim U\left(0.01, 0.99\right)\) and shape
parameter \(\beta \sim U\left(1.00, 5.00 \right)\). Location and scale
parameters were set as \(\mu = 0\) and \(\sigma = 1\), respectively.
\(n = \lceil 10^\gamma \rceil\) instances were then randomly drawn, with
\(\gamma \sim U\left(2, 4\right)\), i.e., between \(100\) and \(10000\)
values are drawn to create \(\mathbf{X}_i\).

Outlier values were then drawn to randomly replace a fraction of the
values of \(\mathbf{X}_i\). This was repeated \(m_{\text{out}} = 10\)
times, with outlier fractions regularly spaced in \([0.00, 0.10]\). Thus
up to 10 percent of the values could be replaced by outliers. Outlier
values were set according to John W. Tukey (1977), as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

\begin{equation}
x_{out} =
\begin{cases}
Q_1 - \left(1.5 - x^{*} \right) \text{IQR} & \text{if } x^{*} < 0 \\
Q_3 + \left(1.5 + x^{*} \right) \text{IQR} & \text{if } x^{*} \geq 0
\end{cases}
\end{equation}

\(Q_1\), \(Q_3\) and \(\text{IQR}\) are the first quartile, third
quartile and interquartile range of \(\mathbf{X}_i\), respectively.
Outlier values randomly replaced values in \(\mathbf{X}_i\) to create
\(\mathbf{X}_{i,j}\), with
\(j \in \left\{ 1, \ldots, m_{\text{out}} \right\}\)

To find the optimal values for the weighting function parameters
\(\delta_1\) and \(\delta_2\) (if applicable), we minimised the absolute
difference between the \(\lambda_{r}\) parameter obtained for robust
transformation in the presence of outliers, and the \(\lambda_0\)
parameter obtained using the non-robust transformations in absence of
outliers:

\begin{equation}
\label{eqn:minimisation-weighting}
\left\{ \hat{\delta}_1, \hat{\delta}_2 \right\} = \argmin_{\delta_1, \delta_2} \sum_{i=1}^{m_d} \sum_{j=1}^{m_{\text{out}}} \left| \lambda_r \left(\mathbf{X}_{i, j}; \delta_1, \delta_2 \right) - \lambda_0 \left(\mathbf{X}_i \right) \right|
\end{equation}
